From Causal Discovery to Intervention Simulation: Modeling Precarity and Depression in the HELIUS Study

### 2025-05-20 draft_v1/draft_v1.tex
draft update

\date{2025-05-18}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates the dynamic relationship between
precarious life conditions and depression by combining constraint-based
causal discovery with computational modeling. Using cross-sectional data
from the HELIUS study, we applied cycle-capable causal discovery
algorithms to uncover directional relationships between financial
stress, domain-specific precariousness indicators, and depressive
symptoms. Our multi-resolution analysis---spanning both aggregate and
symptom-level constructs---identified financial stress as a robust
driver of depression, and highlighted key symptoms, such as sleep
disturbance and depressed mood, as likely initiators or reinforcers of
downstream precariousness. To explore how these structural insights
shape intervention effectiveness, we developed a simplified nonlinear
dynamical model simulating the co-evolution of depression and social
precarity under varying levels of financial stress reduction. Simulation
results revealed that the system's responsiveness depends not only on
the intensity of external intervention but also on internal structural
features such as feedback strength and noise. In particular, some
configurations exhibited bistability, requiring interventions to surpass
a critical threshold to trigger lasting improvement, while others
responded more proportionally. Together, these findings emphasize that
internal feedback dynamics---not just external adversity---can shape the
impact of mental health interventions. More broadly, our study offers a
scalable framework for integrating causal discovery and dynamical
simulation to better understand and inform policy responses to complex
psychosocial phenomena.
\end{abstract}

@@

This study aims to investigate the causal mechanisms linking
precariousness and depression---and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precarity--depression relationship
affect the outcomes of stress-reducing interventions?} To answer this,
we combine two approaches. First, we use cycle-capable causal discovery
algorithms to infer directional relationships among financial stress,
domain-specific precariousness factors, and depressive symptoms. By
analyzing both aggregate scores (e.g., PHQ-9 total, composite precarity
index) and disaggregated variables (individual symptoms and life-domain
indicators), we identify both broad associations and fine-grained causal
pathways. Second, we translate these insights into a computational
dynamical model that simulates how depression and precarity co-evolve
under varying levels of external stress. This model enables us to probe
how system properties---such as feedback strength and stochastic
noise---govern responsiveness to intervention. In some configurations,
the system exhibits bistability, remaining stuck in a high-risk state
unless external support exceeds a critical threshold. In others, it
shows unistable behavior, responding more gradually to incremental
changes. By integrating causal discovery with dynamic simulation, we aim
to demonstrate how the internal structure of the depression--precarity
system shapes its susceptibility to change---and, by extension, how
intervention strategies might be tailored to the architecture of
underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study,
which has been described in detail elsewhere (Snijder et al., 2017;
Stronks et al., 2013). It captures the diverse population of the city of
Amsterdam by including the six main ethnic groups and which provides
comprehensive health and lifestyle data, including depressive symptoms
as measured by the PHQ-9 (Galenkamp et al., 2017). To operationalize
indicators of precariousness, we draw on the framework outlined in
previous research (Elsenburg et al., 2025) and select a set of relevant
variables. To ensure a robust representation of precariousness in our
causal discovery models, we conducted various exploratory analyses to
identify consistent and meaningful data-driven factor structures. Based
on these analyses, we identified five precariousness factors, including
two related to recent stressors, each comprising multiple variables as
outlined below. DFurther details on the HELIUS study and factor
extraction are provided in the \hyperref[sec-appendix]{Appendix}.

\begin{itemize}
\tightlist
\item
  Employment precariousness: \texttt{emp\_stat}, \texttt{work\_sit}.
\item
  Social precariousness: \texttt{soc\_freq}, \texttt{soc\_adq}.
\item
  Housing precariousness: \texttt{nb\_safe}, \texttt{nb\_res},
  \texttt{nb\_rent}, \texttt{cul\_rec}.
\item
  Recent relational stressors: \texttt{frd\_brk12}, \texttt{conf12}.
\item
  Recent financial stressors: \texttt{fincri12}, \texttt{inc\_diff}.
\end{itemize}

After preprocessing, the HELIUS dataset comprises 21,628 samples. Along
with the five precariousness factors, we also compute an overall
precarity score as the combined value of these factors. Additionally,
PHQ-9 scores are included to represent depression, both as a total sum
score and as individual symptom scores. In the subsequent causal
discovery analysis, we examine the relationship between depression and
precarity using both aggregated sum scores and their individual-level
representations. Refer to Figure~\ref{fig-dist} for the overall
distributions of the variables used in the analysis.

@@

To uncover directional relationships between precariousness and
depression, we used constraint-based causal discovery algorithms that
allow for feedback loops and latent confounding. Specifically, we
applied the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
Inference} (CCI) algorithms, both of which can detect cycles and
non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
reference, we also included the \emph{PC} algorithm, one of the most
widely used methods for acyclic causal discovery (Spirtes et al., 2001).

A key challenge in applying these algorithms to the HELIUS dataset is
that many variables are non-Gaussian and their relationships are likely
nonlinear. To address this, we supplemented the commonly used Gaussian
conditional independence (CI) test---based on partial
correlations---with a non-parametric alternative: the \emph{Randomized
Conditional Correlation Test} (RCoT). RCoT approximates kernel-based CI
testing using random Fourier features, significantly reducing
computational demands while maintaining sensitivity to nonlinear
dependencies (Strobl et al., 2019; Zhang et al., 2012).

We examined the causal structure using three complementary approaches:

@@

To ensure robust estimation and guard against sensitivity to parameter
settings, we implemented a systematic grid of analysis conditions:

\begin{itemize}
\tightlist
\item
  Significance levels: \(\alpha\) = 0.01 and 0.05
\item
  Bootstrap thresholds: 0.5, 0.6, 0.7, 0.8
\item
  CI tests: Gaussian and RCoT
\item
  Algorithms: FCI, CCI, and PC
\end{itemize}

Each combination was bootstrapped 100 times for sum-score level analyses
and 30 times for symptom-level analyses, yielding a total of 1,600
graphs per algorithm for the full analysis. For symptom networks, we
fixed the skeleton structure using a consensus graph estimated across
PC, FCI, and CCI (\(\alpha\) = 0.01, RCoT). This constraint improves
computational efficiency without limiting the estimation of directional
edges between symptoms and precariousness variables.

To summarize the results, we adopted a two-level aggregation strategy.
First, within each analysis condition, we recorded the most frequently
observed edge-endpoint type (e.g., directed, undirected, uncertain) from
the bootstraps. Then, across all conditions, we tallied the most
frequent endpoint type per edge. In cases of ties, we assigned dashed
edges to reflect uncertainty. For transparency, we accompany each graph
with a table showing the relative frequencies of each edge symbol.

This procedure yields a final graph for each algorithm and analysis
type---offering both robustness and interpretability. See
Figure~\ref{fig-workflow} for an overview of the causal discovery
pipeline.

For detailed explanations of the algorithms, edge interpretations, and
graph types (e.g., PAG, MAAG, CPDAG), as well as algorithmic assumptions
and limitations, see the \hyperref[sec-appendix]{Appendix}.

@@

\includegraphics[width=0.8\textwidth,height=\textheight]{img/simsetup2.png}

@@

\caption{\label{fig-workflow}Analysis workflow applied across all three
algorithms.}

@@

\includegraphics[width=0.35\textwidth,height=\textheight]{img/conceptmodel.png}

@@

\caption{\label{fig-concept}Conceptual high-level causal structure of
precariousness and depression system.}

@@

To identify realistic parameter settings for the simulation model, we
calibrated six free parameters---\(\alpha_1\), \(\alpha_2\),
\(\beta_1\), \(\beta_2\), and noise amplitudes \(\sigma_1\),
\(\sigma_2\)---using the Non-dominated Sorting Genetic Algorithm II
(NSGA-II) (Deb et al., 2002), a multi-objective evolutionary
optimization method well-suited for noisy and high-dimensional search
spaces. The goal was to match simulated system behavior to empirical
summary statistics derived from the HELIUS dataset, including the means
and variances of depression (\(D\)) and social precariousness (\(P\)),
as well as their partial correlations after adjusting for financial
stress (\(S\)), specifically \(\rho_{DP.S}\), \(\rho_{DS.P}\), and
\(\rho_{PS.D}\). Given the stochastic nature of the model, each
candidate parameter set was evaluated by simulating 1,000 independent
trajectories, each representing an individual with a fixed \(S\) value
sampled from the observed distribution. Final-state values were
extracted and aggregated to compute population-level summary statistics.
Fitness was assessed as the absolute deviation from the empirical
targets, aggregated across all objectives. NSGA-II was implemented via
the nsga2R package (Franz \& Nakamura, 2015), with a population of 300
evolved over 300 generations using standard hyperparameters (crossover
probability = 0.9, mutation probability = 0.25, and tournament size =
2). This process yielded a diverse ensemble of Pareto-efficient
solutions balancing trade-offs between objectives. From the resulting
Pareto front (Figure~\ref{fig-pareto}), we selected the top 30 parameter
sets with the lowest combined error to use in the subsequent simulations
assessing intervention effects.

@@

For each of the top 30 calibrated parameter sets, we simulate the
system's evolution across 300 individuals under a continuum of
intervention levels. Each simulation run generates final values for
\(D\) and \(P\), which are then used to assess how internal model
parameters influence the system's behavior under changing external
conditions.

@@

The causal graph in panel (a) highlights key pathways in the
relationships between precarity factors and depression (PHQsum).
Employment precarity (P.emp) and social precarity (P.soc) are not
identified as causes of depression, whereas financial stress (S.fin)
appears to play a causal role. While P.emp and P.soc are generally not
recognized as causes of other precariousness factors, S.fin emerges as a
potential cause, as indicated by its circle edge endpoint. Supporting
this, the proportion matrix plot in panel (b) shows that S.fin has some
probability of causally influencing either P.emp or P.soc.

@@

Relational stress (S.rel) plays a more nuanced role, interacting with
depression and S.fin through a latent confounder. Its relationship with
P.soc remains less certain, as the proportion table suggests a fair
probability that this connection is absent, uncertain, or bidirectional.
Meanwhile, housing precarity (P.hou) is not directly causally related to
depression but is linked to P.emp and S.fin. While P.emp and S.fin are
identified as non-causes of P.hou, it remains unclear whether P.hou
causally influences P.emp or if their relationship is mediated by an
unobserved confounder.

The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
largely consistent with those from the FCI algorithm, with one
exception: CCI does not identify S.fin as a cause of depression.
Additionally, CCI introduces greater uncertainty in edge directions,
with more ties mainly between circle and arrowhead endpoints, and tends
to favor arrowheads more frequently than FCI. Despite these differences,
the skeleton structure and overall causal directions derived from CCI
align well with the results of the FCI algorithm, supporting the key
pathways.

@@

Moving from the sum score representation to the disaggregated
symptom-level graph provides a more granular perspective on the causal
relationships between precariousness factors and depression symptoms.
Unlike the sum score graph, which aggregates all symptoms into a single
measure---potentially obscuring nuanced relationships---the
symptom-level graph highlights heterogeneity in how precariousness
factors influence individual depressive symptoms: con (concentration),
slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal),
anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.

The symptom-level graph in Figure~\ref{fig-sym} panel (a) reveals a more
complex and interconnected structure than the sum score graph,
reflecting the strong interdependence among symptoms and suggesting much
presence of latent confounders, as indicated by numerous bidirectional
edges. As before, dashed edges denote ties, with specific ties marked by
bold horizontal bars. In Figure~\ref{fig-sym} panel (b), the proportion
matrix provides more insight into these directionalities. Here, opaque
green dominates most symptom-to-symptom connections, indicating that
arrowheads are the most frequently inferred edge type. However, blue
regions, which correspond to circle endpoints, are particularly common
for anh, slp, ene, and sui, reflecting uncertainty in the causal
direction among these symptoms. Notably, the only arrowtail connection
appears between dep and anh, suggesting that anhedonia (anh) is a
potential cause of depressed mood (dep).

Looking at the symptom-precariousness connections, one of the key
patterns is that ties are most frequently found in the relationships
between individual symptoms and precariousness factors. This suggests
that these causal links may be less stable across different conditions.
Upon closer examination, most ties arise due to discrepancies between
the Gaussian CI test and the RCoT test---with Gaussian CI favoring
arrowheads and RCoT more often predicting the absence of an edge.

This discrepancy likely arises from a combination of factors. First, as
the analysis shifts from aggregated to disaggregated variables, the
number of conditional independence (CI) tests increases, leading to a
loss of statistical power in high-dimensional causal discovery. With
more variables being conditioned on, the data becomes sparser, making it
harder to detect weaker dependencies. Additionally,
over-conditioning---controlling for too many variables---can
artificially remove statistical associations, leading to false
negatives. Beyond this general statistical power issue, the fundamental
differences in how these CI tests operate contribute to the
inconsistency. RCoT, as a nonparametric test, does not assume linearity
and can capture both linear and nonlinear dependencies. However, it is
more conservative and requires larger sample sizes, making it more prone
to false negatives in high-dimensional settings. In contrast, partial
correlation (Gaussian CI test) assumes linear Gaussian relationships,
making it more permissive and sometimes detecting weak dependencies that
may not be statistically meaningful. As a result, when Gaussian CI
detects an edge while RCoT does not, it could indicate that the
relationship is weak, strictly linear, or requires more data for
reliable nonparametric detection. This explains why many ties in the
proportion matrix show a 50-50 split between symbols---where half of the
conditions with RCoT suggest the absence of an edge, while the other
half using Gaussian CI test favor an arrowhead. This systematic
divergence highlights the challenges of causal discovery in
high-dimensional settings, where varying statistical assumptions can
lead to differing conclusions about the presence and direction of causal
relationships.

Despite these differences, some consistent patterns emerge across both
CI tests, particularly in the case of financial stress (S.fin), which
appears to be connected to nearly all other variables---though many of
these connections are marked by ties, reflecting uncertainty in
directionality. One exception is the stronger evidence suggesting that
S.fin may cause changes in app (appetite), while its relationships with
other symptoms, such as ene, mot, and dep, remain ambiguous. Similarly,
social precarity (P.soc) exhibits numerous connections with depressive
symptoms, with most edges pointing toward P.soc rather than outward from
it. This suggests that depressive symptoms, particularly dep and glt,
may contribute to worsening social precarity rather than the other way
around. Additionally, anh also shows some probability of causally
influencing P.soc, reinforcing the idea that social precarity is more
often a consequence rather than a driver of depressive symptoms.

Other precariousness factors exhibit more uncertain but still notable
relationships. Relational stress (S.rel) shows weak but existing
connections with slp, glt, and app, though directionality remains
unclear in many cases. Employment precarity (P.emp) also connects to
symptoms like slp, mot, and con, but, like S.rel, these relationships
exhibit a 50/50 split in directionality, reflecting uncertainty in the
inferred causal paths. In line with the aggregated graph in
Figure~\ref{fig-sum}, housing precarity (P.hou) does not appear to have
any direct relationship with depression symptoms but consistently shows
associations with employment precarity (P.emp) and financial stress
(S.fin). The causal relationships among precariousness factors remain
largely unchanged from the aggregated analysis, with financial stress
(S.fin) exhibiting the strongest tendency to influence other
precariousness factors.

Among depressive symptoms, dep, glt, and slp appear to be the most
connected to precariousness factors, while P.soc has the most
connections with symptoms, predominantly as a recipient rather than a
driver of influence. Within the symptom network, slp, sui, and anh
emerge as causally influential symptoms, as they exhibit more outgoing
arrows compared to other symptoms. On the other hand, con, mot, dep, and
glt, despite having high connectivity, predominantly receive incoming
arrows, indicating they are more likely effects rather than causes.
Considering both symptom-precariousness connections and symptom-level
dynamics, sleep disturbance (slp) emerges as a central symptom, given
its strong ties to precariousness factors and its influential role
within the symptom network. This suggests that sleep issues could be an
initiating symptom, particularly sensitive to relational stress and
employment precarity. On the other hand, social precarity (P.soc) serves
as a key bridge between the depression and precariousness subsystems, as
depressive symptoms appear to feed back into social precarity,
reinforcing a self-sustaining dynamic between depression and precarious
conditions.

>>>>

\date{2025-05-20}
\begin{document}
\maketitle
\begin{abstract}
\noindent This study investigates the dynamic relationship between
precarious life conditions and depression by combining constraint-based
causal discovery with computational modeling. Using cross-sectional data
from the HELIUS study, we applied cycle-capable causal discovery
algorithms to uncover directional relationships between domain-specific
precariousness indicators, and depressive symptoms. Our multi-resolution
analysis---spanning both aggregate and symptom-level
constructs---identified financial stress as a robust driver of
depression, and highlighted key symptoms, such as sleep disturbance and
depressed mood, as likely initiators or reinforcers of downstream
precariousness. To explore how these structural insights shape
intervention effectiveness, we developed a simplified nonlinear
dynamical model simulating the co-evolution of depression and social
precarity under varying levels of financial stress reduction. Simulation
results revealed that the system's responsiveness depends not only on
the intensity of external intervention but also on internal structural
features such as feedback strength and noise. In particular, some
configurations exhibited bistability, requiring interventions to surpass
a critical threshold to trigger lasting improvement, while others
responded more proportionally. Together, these findings emphasize that
internal feedback dynamics---not just external adversity---can shape the
impact of mental health interventions. More broadly, our study offers a
scalable framework for integrating causal discovery and dynamical
simulation to better understand and inform policy responses to complex
psychosocial phenomena.
\end{abstract}

@@

This study aims to investigate the causal mechanisms linking
precariousness and depression---and how those mechanisms shape system
responses to external intervention. Specifically, we ask: \emph{How does
the internal configuration of the precarity--depression relationship
affect the outcomes of stress-reducing interventions?} To answer this,
we combine two approaches. First, we use cycle-capable causal discovery
algorithms to infer directional relationships among domain-specific
precariousness factors, and depressive symptoms. By examining both
aggregated measures (depression sum score and a composite precarity
index) and disaggregated variables (individual depressive symptoms and
specific life-domain indicators), we capture broad patterns of
relationship as well as fine-grained causal pathways. Second, we
translate these insights into a computational dynamical model that
simulates how depression and precarity co-evolve under varying levels of
external stress. This model enables us to probe how system
properties---such as feedback strength and stochastic noise---govern
responsiveness to intervention. In some configurations, the system
exhibits bistability, remaining stuck in a high-risk state unless
external support exceeds a critical threshold. In others, it shows
unistable behavior, responding more gradually to incremental changes. By
integrating causal discovery with dynamic simulation, we aim to
demonstrate how the internal structure of the depression--precarity
system shapes its susceptibility to change---and, by extension, how
intervention strategies might be tailored to the architecture of
underlying vulnerability.

@@

We use data from the HELIUS (HEalthy LIfe in an Urban Setting) study,
which has been described in detail elsewhere (Snijder et al., 2017;
Stronks et al., 2013). It captures the diverse population of the city of
Amsterdam by including the six main ethnic groups and which provides
comprehensive health and lifestyle data, including depressive symptoms
as measured by the PHQ-9 (Galenkamp et al., 2017). To operationalize
indicators of precariousness, we draw on the framework outlined in
previous research (Elsenburg et al., 2025) and select a set of relevant
variables. To ensure a robust representation of precariousness in our
causal discovery models, we conducted exploratory analyses to identify
consistent and meaningful data-driven factor structures. These analyses
led to the identification of five distinct precariousness
factors---three domain-based and two reflecting recent stressors---each
composed of multiple observed variables, as outlined below. Additional
details on the HELIUS study and the factor extraction process are
provided in the \hyperref[sec-appendix]{Appendix}.

\begin{itemize}
\tightlist
\item
  Employment precariousness: \texttt{emp\_stat} (current employment
  status), \texttt{work\_sit} (nature of work situation).
\item
  Social precariousness: \texttt{soc\_freq} (frequency of social
  contact), \texttt{soc\_adq} (perceived adequacy of social support).
\item
  Housing precariousness: \texttt{nb\_safe} (neighborhood safety),
  \texttt{nb\_res} (access to neighborhood resources), \texttt{nb\_rent}
  ( proportion of rental housing), \texttt{cul\_rec} (availability of
  cultural resources).
\item
  Recent relational stressors: \texttt{frd\_brk12} (experience of a
  friendship breakup), \texttt{conf12} (recent interpersonal or
  relational conflict).
\item
  Recent financial stressors: \texttt{fincri12} (experience of a
  financial crisis), \texttt{inc\_diff} ( difficulty managing household
  income).
\end{itemize}

After preprocessing, the HELIUS dataset includes 21,628 samples. In
addition to the five identified precariousness factors, we computed an
overall precarity score by aggregating their values. Depression is
represented using the PHQ-9 (Kroenke et al., 2001), both as a total sum
score and as individual symptom scores. In the subsequent causal
discovery analysis, we explore the relationship between depression and
precarity using both aggregated scores and their disaggregated
components. See Figure~\ref{fig-dist} for the overall distributions of
the variables used in the analysis.

@@

To uncover directional relationships between precariousness and
depression, we use constraint-based causal discovery algorithms that
allow for feedback loops and latent confounding. Specifically, we apply
the \emph{Fast Causal Inference} (FCI) and \emph{Cyclic Causal
Inference} (CCI) algorithms, both of which can detect cycles and
non-acyclic structures (Mooij \& Claassen, 2020; Strobl, 2019). For
reference, we also include the \emph{PC} algorithm, one of the most
widely used methods for acyclic causal discovery (Spirtes et al., 2001).
While our full analysis includes results from three causal discovery
algorithms---FCI, CCI, and PC---we focus on FCI in the main text, as it
offers a strong balance between flexibility and theoretical soundness.
FCI accommodates latent confounding and potential feedback, making it
well-suited for the complexity of our data (Mooij \& Claassen, 2020).
The PC algorithm is included for reference but relies on more
restrictive assumptions, such as acyclicity and no unmeasured
confounding. CCI, while capable of modeling cycles explicitly, comes
with certain theoretical limitations. Results from both PC and CCI are
provided in the \hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-cci}, Section~\ref{sec-pc}) for completeness and
comparison. For readers interested in detailed explanations of the
algorithms, edge interpretations, and graph types (e.g., PAG, MAAG,
CPDAG), as well as algorithmic assumptions and limitations, please refer
to the \hyperref[sec-appendix]{Appendix}
(Section~\ref{sec-causalprimer}).

A key challenge in applying these algorithms to the HELIUS dataset is
that many variables are non-Gaussian and likely exhibit nonlinear
relationships. While kernel-based conditional independence tests (such
as KCIT) are well-suited for capturing such complex dependencies, they
are computationally intensive---scaling quadratically with sample size
due to the need to invert large kernel matrices---making them
impractical for large datasets like HELIUS (Rahimi \& Recht, 2007). To
address this, we use the Randomized Conditional Correlation Test (RCoT),
a nonparametric alternative that approximates kernel methods using
random Fourier features. This reduces computational complexity from
quadratic to linear in sample size, substantially lowering runtime while
preserving sensitivity to nonlinear relationships (Strobl et al., 2019;
Zhang et al., 2012). For further details, see Section~\ref{sec-rcot}.

We examine the causal structure using three complementary approaches:

@@

To robustly estimate the causal structure between depression and
precariousness variables, we implement a comprehensive bootstrapped
causal discovery procedure. This procedure systematically varies key
analysis parameters across a grid of configurations, enabling us to
assess the stability of inferred relationships under different settings.
Specifically, we vary four main components:

\begin{itemize}
\tightlist
\item
  Significance levels for conditional independence tests: \(\alpha\) =
  0.01 and 0.05.
\item
  Stability thresholds: 0.5, 0.6, 0.7, and 0.8, which determine how
  frequently an edge must appear across bootstraps to be retained.
\item
  Conditional independence (CI) tests: Gaussian CI test (partial
  correlation) and RCoT (a nonparametric test).
\item
  Causal discovery algorithms: FCI, CCI, and PC.
\end{itemize}

For each unique parameter combination (2 significance levels × 4
thresholds × 2 CI tests = 16 settings), we run 100 bootstrap samples per
algorithm for the aggregated analyses (see Step 1 in
Figure~\ref{fig-workflow}). For symptom-level analyses, which are more
computationally intensive, we use 30 bootstrap samples per
setting.\footnote{For symptom-level analyses, we further improve
  computational efficiency by fixing the skeleton structure---the
  undirected network of potential connections---using a consensus graph
  derived from all three algorithms (with \(\alpha = 0.01\) and the RCoT
  test). This allows subsequent analyses to focus exclusively on
  estimating edge directions, avoiding the repeated and costly
  computation of skeletons, and thus substantially reducing runtime.}

To summarize these results, we apply a two-level aggregation strategy:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Within each condition} (e.g., fixed \(\alpha\), threshold, CI
  test, and algorithm), we identify the most frequently occurring edge
  type (e.g., arrowhead, circle, none) across bootstraps for each pair
  of variables (Step 2 in Figure~\ref{fig-workflow}).
\item
  \textbf{Across conditions}, we then determine the most dominant edge
  type based on majority voting. In cases where no edge type is clearly
  dominant (i.e., a tie), we mark the edge as uncertain, represented
  with a dashed line in the final graph (Step 3 in
  Figure~\ref{fig-workflow}).
\end{enumerate}

This summarization process is repeated separately for each causal
discovery algorithm (FCI, CCI, PC), yielding a single robust graph per
algorithm. These final graphs represent only the most stable and
consistent relationships across a wide range of plausible
parameterizations, enhancing both reliability and interpretability. To
support transparency, each final graph is accompanied by a matrix
summarizing the relative frequency of each edge type for every variable
pair---enabling readers to assess the degree of uncertainty or agreement
underlying each connection.

@@

\includegraphics[width=0.9\textwidth,height=\textheight]{img/simsetup.pdf}

@@

\caption{\label{fig-workflow}Causal discovery workflow applied across
all three algorithms (FCI, CCI, and PC). \emph{The process involves
three main steps: (1) bootstrapping the dataset to create multiple
resampled datasets; (2) estimating causal graphs under varying parameter
configurations; and (3) summarizing results by identifying the most
frequent edge type across bootstraps and settings to construct a final
stable graph.}}

@@

\includegraphics[width=0.4\textwidth,height=\textheight]{img/conceptmodel.png}

@@

\caption{\label{fig-concept}High-level conceptual model of causal
structure linking social precariousness (P.soc) and depression (PHQsum),
influenced by external financial stress (S.fin).}

@@

To identify realistic parameter settings for the simulation model, we
calibrate six free parameters---\(\alpha_1\), \(\alpha_2\), \(\beta_1\),
\(\beta_2\), and noise amplitudes \(\sigma_1\), \(\sigma_2\)---using the
\emph{Non-dominated Sorting Genetic Algorithm II} (NSGA-II) (Deb et al.,
2002), a multi-objective evolutionary optimization method well-suited
for noisy and high-dimensional search spaces. The goal is to match
simulated system behavior to empirical summary statistics derived from
the HELIUS dataset. Specifically, we aim to match the observed means and
variances of depression (\(D\)) and social precariousness (\(P\)), along
with all three partial correlations among \(D\), \(P\), and financial
stress (\(S\)): \(\rho_{DP.S}\), \(\rho_{DS.P}\), and \(\rho_{PS.D}\).

Given the stochastic nature of the model, each candidate parameter set
is evaluated by simulating 1,000 virtual individuals, each assigned a
fixed \(S\) value sampled from the empirical distribution. From these
simulations, we compute population-level summary statistics based on the
final states. The fitness of each parameter set is measured by how
closely its summary statistics matched the empirical data, using the
total absolute deviation across all objectives.

NSGA-II is implemented via the \texttt{nsga2R} package (Franz \&
Nakamura, 2015), with a population of 500 evolved over 300 generations
using standard hyperparameters (crossover probability = 0.9, mutation
probability = 0.25, and tournament size = 2). This process yielded a
diverse ensemble of Pareto-efficient solutions balancing trade-offs
between objectives. From the resulting Pareto front
(Figure~\ref{fig-pareto}), we selected the top 30 parameter sets with
the lowest combined error to use in the subsequent simulations assessing
intervention effects.

@@

For each of the top 30 parameter sets identified through calibration, we
run simulations on 300 virtual individuals across this full range of
intervention levels. In each simulation, we observe how depression
(\(D\)) and precariousness (\(P\)) evolve over time, capturing their
final values after the system stabilizes. This setup allows us to
evaluate how different internal model dynamics (e.g., feedback strength)
influence the system's response to external change.

@@

The causal graph in panel (a) highlights key pathways in the
relationships between precarity factors and depression (\emph{PHQsum}).
Employment precarity (\emph{P.emp}) and social precarity (\emph{P.soc})
are not identified as causes of depression, whereas financial stress
(\emph{S.fin}) appears to play a causal role. While \emph{P.emp} and
\emph{P.soc} are generally not recognized as causes of other
precariousness factors, \emph{S.fin} emerges as a potential cause, as
indicated by its circle edge endpoint. Supporting this, the proportion
matrix plot in panel (b) shows that \emph{S.fin} has some probability of
causally influencing either \emph{P.emp} or \emph{P.soc.}

@@

Relational stress (\emph{S.rel}) plays a more nuanced role, interacting
with depression and \emph{S.fin} through a latent confounder. Its
connection to \emph{P.soc} is less definitive---the proportion matrix in
Figure~\ref{fig-sum-2} indicates a considerable probability that this
relationship is either absent, ambiguous, or potentially bidirectional.
Meanwhile, housing precarity (\emph{P.hou}) is not directly causally
related to depression but is linked to \emph{P.emp} and \emph{S.fin}.
While \emph{P.emp} and \emph{S.fin} are identified as non-causes of
\emph{P.hou}, it remains unclear whether \emph{P.hou} causally
influences \emph{P.emp} or if their relationship is mediated by an
unobserved confounder.

The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
largely consistent with those from the FCI algorithm, with one
exception: CCI does not identify \emph{S.fin} as a cause of depression.
Additionally, CCI introduces greater uncertainty in edge directions,
with more ties mainly between circle and arrowhead endpoints, and tends
to favor arrowheads more frequently than FCI. Despite these differences,
the skeleton structure and overall causal directions derived from CCI
align well with the results of the FCI algorithm, supporting the key
pathways.

@@

Moving from the sum score representation to the disaggregated
symptom-level graph provides a more granular perspective on the causal
relationships between precariousness factors and depression symptoms.
Unlike the sum score graph, which aggregates all symptoms into a single
measure---potentially obscuring nuanced relationships---the
symptom-level graph reveals the heterogeneity in how specific symptoms
(\emph{con} (concentration), \emph{slp} (sleep), \emph{ene} (energy),
\emph{app} (appetite), \emph{mot} (motor), \emph{sui} (suicidal),
\emph{anh} (anhedonia), \emph{glt} (guilt), and \emph{dep} (depressed
mood)) are influenced by, and in turn influence, different forms of
precariousness.

The symptom-level graph in Figure~\ref{fig-sym-1} reveals a more complex
and interconnected structure than the sum score graph, reflecting the
strong interdependence among symptoms and suggesting much presence of
latent confounders, as indicated by numerous bidirectional edges. As
before, dashed edges denote ties, with specific ties marked by bold
horizontal bars. In Figure~\ref{fig-sym-2}, the proportion matrix
provides more insight into these directionalities. Here, opaque green
dominates most symptom-to-symptom connections, indicating that
arrowheads are the most frequently inferred edge type. However, blue
regions, which correspond to circle endpoints, are particularly common
for \emph{anh}, \emph{slp}, \emph{ene}, and \emph{sui}, reflecting
uncertainty in the causal direction among these symptoms. Notably, the
only arrowtail connection appears between \emph{dep} and \emph{anh},
suggesting that anhedonia is a potential cause of depressed mood.

Looking at the symptom-precariousness connections, one of the key
patterns is that ties are most frequently found in the relationships
between individual symptoms and precariousness factors. This suggests
that these causal links may be less stable across different conditions.
Closer examination reveals that most ties arise due to discrepancies
between the Gaussian CI test and the RCoT test---with Gaussian CI
favoring arrowheads and RCoT more often predicting the absence of an
edge.

This discrepancy likely arises from a combination of factors. As the
analysis moves from aggregated to disaggregated variables, the number of
CI tests increases substantially, leading to a loss of statistical
power. With more variables being conditioned on, the effective sample
size per test diminishes, making it harder to detect subtle
dependencies. Over-conditioning---controlling for too many
variables---can also obscure real relationships, resulting in false
negatives. Beyond these statistical power issues, the underlying
assumptions of the two CI tests contribute to their divergence. RCoT is
a nonparametric test that captures both linear and nonlinear
dependencies without assuming a specific distributional form. However,
it is more conservative and requires larger sample sizes, especially in
high-dimensional settings. In contrast, the Gaussian CI test is based on
partial correlations and assumes linear Gaussian relationships, making
it more permissive and more likely to detect weak or spurious
associations. As a result, it is common to see edges where Gaussian CI
suggests a directional relationship, while RCoT indicates no edge. In
the proportion matrix, this appears as a 50--50 tie between absence and
arrowhead symbols. This systematic divergence illustrates the broader
challenge of causal discovery in high-dimensional data: different
statistical assumptions can lead to different conclusions about
whether---and in which direction---variables are causally related.

Despite these differences, some consistent patterns emerge across both
CI tests, particularly in the case of \emph{S.fin}, which appears to be
connected to nearly all other variables---though many of these
connections are marked by ties, reflecting uncertainty in
directionality. One exception is the stronger evidence suggesting that
\emph{S.fin} may cause changes in \emph{app}, while its relationships
with other symptoms, such as \emph{ene}, \emph{mot}, and \emph{dep},
remain ambiguous. Similarly, \emph{P.soc} exhibits numerous connections
with depressive symptoms, with most edges pointing toward \emph{P.soc}
rather than outward from it. This suggests that depressive symptoms,
particularly \emph{dep} and \emph{glt}, may contribute to worsening
social precarity rather than the other way around. Additionally,
\emph{anh} also shows some probability of causally influencing
\emph{P.soc}, reinforcing the idea that social precarity is more often a
consequence rather than a driver of depressive symptoms.

Other precariousness factors exhibit more uncertain but still notable
relationships. \emph{S.rel} shows weak but existing connections with
\emph{slp}, \emph{glt}, and \emph{app}, though directionality remains
unclear in many cases. \emph{P.emp} also connects to symptoms like
\emph{slp}, \emph{mot}, and \emph{con}, but, like \emph{S.rel}, these
relationships exhibit a 50/50 split in directionality, reflecting
uncertainty in the inferred causal paths. In line with the aggregated
graph in Figure~\ref{fig-sum}, \emph{P.hou} does not appear to have any
direct relationship with depression symptoms but consistently shows
associations with \emph{P.emp} and \emph{S.fin}. The causal
relationships among precariousness factors remain largely unchanged from
the aggregated analysis, with \emph{S.fin} exhibiting the strongest
tendency to influence other precariousness factors.
